---
title: "Lab_1"
author: "Yaqi Shi, 1003813180"
output: pdf_document
date: "2023-01-12"
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  out.width = "100%",
  fig.width = 10,
  fig.height = 6.7, 
  fig.retina = 3,
  cache = FALSE)
```


```{r, echo=FALSE}
library(tidyverse)
library(knitr)
library(ggplot2)
```


# Lab Exercises

## Q1 

1. Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme 

```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")

dm1 <- dm %>%
  filter(Age==10|Age==20|Age==30|Age==40) %>%
  mutate(mf_ratio = Male/Female)

dm1 %>% 
  ggplot(aes(x = Year, y = mf_ratio, color = Age)) + 
  geom_line() + 
  scale_color_brewer(palette = "Set1") +
  labs(title = "Ratio of male to female mortality rates over time",
       subtitle = "Age 10, 20, 30 and 40",
       y = "Mortality Ratio")
```

$\;$

## Q2 

2. Find the age that has the highest female mortality rate each year 

```{r}
dm %>%
  select(Year, Age, Female) %>%
  group_by(Year) %>%
  filter(Female==max(Female, na.rm=TRUE)) %>%
  select(-Female)

```

$\;$

## Q3 

3. Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations. 

```{r}
dm %>% 
  mutate(Age = as.numeric(Age)) %>%
  group_by(Age) %>%
  summarize(across(c("Male","Female", "Total"), sd, na.rm = TRUE)) 

```

$\;$

## Q4 

4. The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. 

```{r}
# Handle population data
hmd <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")
hmd <- rename(hmd, Female_Pop  = Female, Male_Pop = Male)
hmd <- hmd %>% select(-Total)

# Handle mortality data
dm <- rename(dm, Female_Mort  = Female, Male_Mort = Male)
dm <- dm %>% select(-Total)

# Combine datasets and calculate weighted mortality rate
comb <- dm %>% left_join(hmd)
comb <- comb %>% mutate(Female_death = Female_Mort * Female_Pop, Male_death = Male_Mort * Male_Pop)

comb <- comb %>% 
  group_by(Year) %>%
  summarize(sum_Female_Pop = sum(Female_Pop, na.rm = TRUE),
            sum_Female_Death = sum(Female_death, na.rm = TRUE),
            sum_Male_Pop = sum(Male_Pop, na.rm = TRUE),
            sum_Male_Death = sum(Male_death, na.rm = TRUE))

comb <- comb %>%
  group_by(Year) %>%
  mutate(Male = sum_Male_Death/sum_Male_Pop,
         Female = sum_Female_Death/sum_Female_Pop) %>%
  select(Year,Male,Female) %>%
  pivot_longer(Male:Female, names_to = "Sex", values_to = "Mortality")

# Produce plots
comb %>% 
  ggplot(aes(x = Year, y = Mortality, color = Sex)) + 
  geom_line() + 
  scale_color_brewer(palette = "Set2") +
  labs(title = "Population weighted average mortality for males and females over time",
       y = "Mortality Rate")
```

Comment: The mortality rate for male were in general higher than the mortality rate for female across all those years and both mortality rates have dropped over time. 




















