---
title: "Lab 4 - Part 2"
author: "Yaqi Shi, 1003813180"
output: pdf_document
date: "2023-02-04"
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  out.width = "100%",
  fig.width = 10,
  fig.height = 6.7, 
  fig.retina = 3,
  cache = FALSE)
```

```{r, echo=FALSE}
library(tidyverse)
library(stringr)
library(janitor)
library(lubridate)
library(ggrepel)
library(tidyverse)
library(knitr)
library(ggplot2)
library(rvest)
```

This file contains the solution for Question 4 only. 

# Question 4: Install rstan and brms

We will be using the packages `rstan` and `brms` from next week. Please install these. Here's some instructions:

- https://github.com/paul-buerkner/brms
- https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started

In most cases it will be straightforward and may not need much more than `install.packages()`, but you might run into issues. Every Stan update seems to cause problems for different OS.

To make sure it works, run the following code:

```{r, echo=FALSE}
library(brms)

x <- rnorm(100)
y <- 1 + 2*x + rnorm(100)
d <- tibble(x = x, y= y)

mod <- brm(y~x, data = d)
summary(mod)

```
















